---
title: "Investigation of the Central Limit Theorem Using the Exponential Distribution"
author: "Alex Van Russelt"
date: "17 December 2017"
header-includes:
  - \usepackage[sfdefault]{roboto}
  - \renewcommand{\familydefault}{\sfdefault}
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Overview
This document investigates the central limit theorem (CLT) through use of an exponential distribution. Random variables are generated using such a distribution, and then averaged. This is repeated a number of times, and it is apparent that the resulting distribution can be well approximated by a normal distribution, as predicted by the CLT.

### Simulations
```{r}
set.seed(123)
simulation_count <- 1000 # The number of times to repeat the experiment
exponential_count <- 40 # The number of exponentials to average
lambda <- 0.2

means <- numeric(length=simulation_count)

for (i in 1:simulation_count){
  data = rexp(exponential_count, lambda)
  means[i] <- mean(data)
}

df_means <- data.frame(means)
```

The `rexp` function is used to generate 40 random variables using the exponential distribution with rate paramater `lambda` set equal to `0.2`. The mean is recorded. This is repeated 1000 times.

### Comparison of sample mean to the theoretical mean
```{r}
sample_mean <- mean(means)
message(paste("Theoretical mean:", format(1/lambda, nsmall=2, digits=2)))
message(paste("Sample mean:", format(sample_mean, nsmall=2, digits=2)))
```

The theoretical mean of an exponential distribution with rate parameter $\lambda$ is given by $\frac{1}{\lambda}$. In this case, $\lambda=0.2$, so the expected mean is 5.00. The sample mean is 5.01, which is very close to the theoretical value.

### Variance of the sample compared to theoretical variance
```{r}
sample_variance <- var(means)
standard_error <- (1 / lambda) ^ 2 / exponential_count
message(paste("Theoretical variance:", format(standard_error, nsmall=2, digits=2)))
message(paste("Sample variance:", format(sample_variance, nsmall=2, digits=2)))
```

The theoretical variance is given by the squared standard error of the means:
$$\sigma_{\bar{x}}^2 = \frac{\sigma^2}{n}$$
where:

1. $\sigma_{\bar{x}}$ is the standard error of the mean.
2. $\sigma$ is the standard deviation of the population (which for an exponential distribution is equal to $\frac{1}{\lambda}$).
3. $n$ is the number of observations (in this case, 40).

This produces a theoretical variance of 0.62. The sample mean is 0.60, which is very close to the theoretical value.

### Comparing distribution of means to a normal distribution
```{r}
qqnorm(means, col="#999999", pch=19,
       main='Quantile plot of sample means')
qqline(means, lwd=2, lty='dashed')
```

Here, a quantile plot is used to compare the quantiles of the means to that of a normal distribution. All data points lie close to the reference line, therefore, the distribution of the means is approximately normal.
