---
title: "Pitch presentation"
author: "Alex Van Russelt"
date: "29 December 2017"
output:
  slidy_presentation: default
  ioslides_presentation: default
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

This shiny web app is intended to allow a user to visualise the relationship between fuel effiency and other variables present in the mtcars dataset. 

The user is able to:

* Select which variable to plot on the *x*-axis using a dropdown box.
* Choose whether to visualise data from vehicles with manual tranmissions or from those with automatic tranmissions.

## Introduction to the `mtcars` dataset
The following information is taken from the `mtcars` help file:

>The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).

To confirm this is the case, we may check that there are 11 features and 32 observations:
```{r, echo=TRUE}
data(mtcars)
message("Total number of features:", ncol(mtcars))
message("Total number of observations:", nrow(mtcars))
```


## Histogram of fuel efficiencies

This plot shows the distribution of fuel efficiencies present in the `mtcars` dataset:

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
ggplot(data=mtcars, aes(x=mpg)) + geom_histogram(binwidth=2, fill='red')
```


##Available *x* variables

The following variables are available to plot on the *x* axis:

| Field  | Detail                 |
| -----  | ------                 |
| `cyl`  | Number of cylinders    |
| `disp` | Displacement (cu.in.)  |
| `hp`   | Gross horsepower       |
| `drat` | Rear axle ratio        |
| `wt`   | Weight (lb/1000)       |
| `qsec` | 1/4 mile time          |
| `vs`   | V/S                    |
| `gear` | Number of forward gears|
| `carb` | Number of carburetors  |