---
title: "Impact of Severe Weather Events in the US Between 1950 and 2011"
author: "Alex Van Russelt"
date: "8 December 2017"
output:
  html_document:
    keep_md: yes
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis
***

This document considers the impact of severe weather events in the US between 1950 and November 2011, as recorded in the NOAA Storm Database. Specifically, it investigates which weather events resulted in the highest number of fatalities, injuries, property damage, and crop damage. It was found that TORNADO events resulted in the most fatalities and injuries, whereas FLOOD and DROUGHT events resulted in the most damage to property and crops respectively.


## Data processing
***

Load libraries:
```{r load_libraries, message=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)
```


Load data from compressed `.csv` file:
```{r load_data}
df <- read.csv(bzfile('StormData.csv.bz2'), stringsAsFactors=FALSE)
```

Some rows have multiple events listed under `EVTYPE`, separated by a `/` character. It is assumed that the first event is the most significant of the two, so here the rows are mutated to remove any additional events:
```{r trim_event_types}
df$EVTYPE <- sub("\\/.*", "", df$EVTYPE)
```

Multipliers for the `PROPDMG` and `CROPDMG` columns are stored as alphabetic characters in the `PROPDMGEXP` and `CROPDMGEXP` columns, these are applied here to create the `PROPDMGCOST` and `CROPDMGCOST` columns:
```{r apply_cost_multiplier}
get_multiplier <- function(multiplier_letter){
  if (multiplier_letter=='K') return(1e3)
  if (multiplier_letter=='M') return(1e6)
  if (multiplier_letter=='B') return(1e9)
  return(1)
}

df$PROPMULTIPLIER <- sapply(df$PROPDMGEXP, FUN=get_multiplier)
df$CROPMULTIPLIER <- sapply(df$CROPDMGEXP, FUN=get_multiplier)

df <- mutate(df, PROPDMGCOST=PROPDMG*PROPMULTIPLIER, CROPDMGCOST=CROPDMG*CROPMULTIPLIER)
```


Prepare summarised data frames containing totalled damages for each event:
```{r summarise_dataframes}
summary_df <- df %>%
              group_by(EVTYPE) %>%
              summarise(fatal=sum(FATALITIES),
                        injured=sum(INJURIES),
                        propdmg=sum(PROPDMGCOST),
                        cropdmg=sum(CROPDMGCOST))

# Put economic data in long-form for plotting later on:
summary_df_cost <- summary_df %>%
                   mutate(total_cost=propdmg+cropdmg) %>% 
                   select(-injured, -fatal) %>%
                   gather(damage_type, cost_of_damage, propdmg, cropdmg)
```



## Results
***

### Part 1: Population health

The events resulting in the most fatalities were **TORNADO** and **EXCESSIVE HEAT**:
```{r fatality_results}
fatality_df <- summary_df %>% 
               arrange(-fatal) %>%
               head() %>%
               select(EVTYPE, fatal)

kable(fatality_df)

g <- ggplot(fatality_df, aes(x=reorder(EVTYPE, -fatal), y=fatal))
g + geom_col() + labs(x='Event type', y='Total fatalities')
```

The events resulting in the most injuries were **TORNADO** and **TSTM WIND**:
```{r injury_results}
summary_df %>% arrange(-injured) %>%
               head() %>%
               select(EVTYPE, injured) %>% 
               kable()
```

### Part 2: Economic cost

The events resulting in the most damage to property were **FLOOD** and **HURRICANE**:
```{r property_results}
summary_df %>% arrange(-propdmg) %>%
               head() %>%
               select(EVTYPE, propdmg) %>% 
               kable()
```

The events resulting in the most damage to crops were **DROUGHT** and **FLOOD**:
```{r crop_results}
summary_df %>% arrange(-cropdmg) %>%
               head() %>%
               select(EVTYPE, cropdmg) %>% 
               kable()
```

Plotting the combined costs, it would appear that the cost of property damage tends to far exceed that of crop damage:
```{r combined_cost_results}
combined_cost_df <- summary_df_cost %>% 
                    arrange(-total_cost) %>% 
                    head(10) %>%
                    mutate(cost_of_damage=cost_of_damage/1e9)

g <- ggplot(combined_cost_df, aes(x=reorder(EVTYPE, -total_cost), y=cost_of_damage, fill=damage_type))
g + geom_col() + labs(x='Event type', y='Cost of damage (USD, billions)') +
    scale_fill_discrete("Damage type", labels=c("Crop", "Property"))
```

